Biomass predictions and VOCC

### 2019-08-09 analysis/VOCC/4-vocc-by-spp.Rmd
streamline merge of vocc and density using lapply (required subsequent code modifications still incomplete)

climate <- "temperature"
temp_threshold <- c(0.75)
optimal_temp <- 6.5 # yellow-eye 4

@@

#climate <- "do and temperature"
max_thresholds <- c(Inf, 0.75)
min_thresholds <- c(0.25, Inf)
critical_thresholds <- c(-0.25, 0.75)
optimal_value <- c(2, 6.5)
```

@@

if (climate == "temperature") {
  variable_names <- c("temp")
  min_thresholds <- c(Inf)
  delta_threshold <- temp_threshold
  optimal_value <- optimal_temp

@@

if (climate == "do") {
  variable_names <- c("do_est")
  max_thresholds <- c(Inf)
  delta_threshold <- -1*(do_threshold)
  optimal_value <- optimal_do

@@

optimal_string <- paste0(optimal_value, collapse = "-")
climate_string <- gsub(" ", "-", gsub("\\/", "-", tolower(climate)))
```

@@

vocc_plot <- gridExtra::grid.arrange(
  vocc_plot_temp, vocc_plot_do,
  ncol = 2,
  top = grid::textGrob(paste("do of ", optimal_value[1], "less", min_thresholds[1],"or temperature of ", optimal_value[2], "plus", max_thresholds[2], " over ", start_time, "-", start_time + delta_t_total, ""))
)

@@

# Extract biomass values to match spatial and temporal range of climate vectors
```{r}
vocc_by_sp <- list()
for (i in seq_along(biomass_list)){

biomass_change <- make_trend_data(biomass_list[[i]],
  ssid = model_ssid,
  start_time = start_time,
  skip_time = skip_time,
  end_time = end_time,
  input_cell_size = input_cell_size,
  scale_fac = scale_fac,
  delta_t_total = delta_t_total,
  delta_t_step = delta_t_step,
  indices = indices,
  variable_names = "est_exp"
)

@@

end <- biomass_change %>% 
  rename(start_density = est_exp_s, 
         end_density = est_exp_e, 
         target_X = x, target_Y = y) %>% 
  select(icell, start_density, end_density, target_X, target_Y)

@@

# if (length(variable_names) == 1) {
#   
#   vectors <- vocc %>%
#     mutate(
#       start_cell = icell, 
#       vector_id = paste0(id, "_to_", tid),
#       initial_value = UQ(rlang::sym(paste0(variable_names, "_s"))), #var_1_s, 
#       source_end_value = UQ(rlang::sym(paste0(variable_names, "_e"))), #var_1_e, 
#       target_value = as.numeric(target_values)
#     ) %>%
#     filter(distance > 1)  %>%
#     select(vector_id, start_cell, x, y, target_X, target_Y, distance, 
#       initial_value, source_end_value, target_value, 
#       km_per_decade, timespan) 
#   
#   if (climate == "do") vectors <- vectors %>% 
#       filter(source_end_value < (optimal_do - do_threshold))
#   if (climate == "temperature") vectors <- vectors %>% 
#       filter(source_end_value > (optimal_temp + temp_threshold))
# 
# } else { # number of variables > 1
  # 
  # vectors <- vocc %>%
  #   mutate(
  #     start_cell = icell, vector_id = paste0(id, "_to_", tid),
  #     initial_value = s, source_end_value = e, target_value = target_values
  #   ) %>%
  #   filter(distance > 1) %>%
  #   select(vector_id, start_cell, x, y, target_X, target_Y, distance, 
  #     initial_value, source_end_value, target_value, 
  #     km_per_decade, timespan) 
  # 
  # # split source_end_value into multiple vectors
  # source_end <- purrr::map_df(strsplit(vectors$source_end_value, " "), 
  #                  function(x) data.frame(t(as.numeric(x))))
  # 
  # vectors <- cbind(vectors, source_end) 
  # 
  # for (j in seq_along(variable_names)){
  #   
  #   source_end_climatej <- paste0("X",j)
  #   vectors <- vectors %>% filter( 
  #     (!!sym(source_end_climatej)) < (optimal_value[j] + critical_thresholds[j])
  #     )
  # }
#  
#   #climate <- "do and temperature"
# # max_thresholds <- c(Inf, 0.75)
# # min_thresholds <- c(0.25, Inf)
# # critical_thresholds <- c(-0.25, 0.75)
# # optimal_value <- c(2, 6.5)
# #

@@

  vectors <- vocc %>%
    mutate(
      start_cell = icell,
      vector_id = paste0(id, "_to_", tid),
      source_end_value = e, # may need to turn into numeric vector later...
      target_value = target_values
    ) %>% filter(distance > 1)

      if (min_thresholds[j] != Inf) { 
        vectors <- filter(vectors, UQ(rlang::sym(paste0(variable_names[j], "_s"))) < optimal_value[j] - min_thresholds[j]) 
        }
      if (max_thresholds[j] != Inf) {
        vectors <- filter(vectors, UQ(rlang::sym(paste0(variable_names[j], "_s"))) > optimal_value[j] + max_thresholds[j])
        }

  #FIXME: decide which variables might cause problems in future merges and remove them...   
  #   vectors <- vectors %>% select(-N) 
  #

}

if (length(vectors[ ,2]) == 0) {
  try({other_coefs <- readRDS("data/vocc-by-spp-sdmTMB-coefs.RDS")})
  coefs <- other_coefs[1,]
  coefs$var <- "NA"
  coefs$beta <- "NA"
  coefs$se <- "NA"
  coefs$species <- spp 
  coefs$lifestage <- "NA"
  coefs$climate <- climate 
  coefs$min_thresholds <- min_string
  coefs$max_thresholds <- max_string
  coefs$optimum <- optimal_string
  coefs$start_time <- start_time
  coefs$timespan <- delta_t_total
  coefs$region <- region
  coefs$ssid_string <- ssid_string

  stop("No vectors.")
}

vector_source <- left_join(vectors, start, by = c("x", "y")) %>% 
  mutate(cell_type = "source", end_climate = source_end_value, source_density = start_density)

vector_target <- left_join(vectors, end, by = c("target_X", "target_Y")) %>% 
  mutate(cell_type = "target", end_climate = target_value)
vector_target <- left_join(vector_target, source_density, by = c("x", "y")) #%>%  
  # filter(end_climate > 2)

# combine source and target cell data
vocc_by_sp[[i]] <- rbind(vector_source, vector_target) %>% 
  filter(!is.na(icell)) %>% 
# remove potential source cells with no fish  
  #FIXME: should their be a flexible min density of fish for a source cell?
  filter(source_density > 0.000001)

# get mean and sd for densities without cell duplication
icells <- vocc_by_sp[[i]] %>% distinct(icell, .keep_all = TRUE)

mean.density <- mean(icells$source_density, na.rm = TRUE)
sd.density <- sd(icells$source_density, na.rm = TRUE)

# calculate standardized density varibles
vocc_by_sp[[i]] <- vocc_by_sp[[i]] %>% 
  mutate(
    perc_change = (end_density - start_density)/start_density, 
    cell_type = as.factor(cell_type),
    #std_change = (perc_change)/sd.change, 
    std_source_density = (source_density - mean.density)/(sd.density*2), 
    mean_change = mean.change, sd_change = sd.change,
    mean_source_density = mean.density, sd_density = sd.density,
# set distance intercept at cell size
    distance = (ifelse(distance > 100, 100, distance)) - input_cell_size, 
    distance2 = distance*distance) %>% 
  rename(X = x, Y = y)

icells <- vocc_by_sp[[i]] %>% distinct(icell, .keep_all = TRUE)

mean.change <- mean(icells$perc_change, na.rm = TRUE)
sd.change <- sd(icells$perc_change, na.rm = TRUE)

vocc_by_sp[[i]] <- vocc_by_sp[[i]] %>% 
  mutate(std_change = (perc_change)/sd.change)

if (length(variable_names) == 1) {
mean.climate <- mean(icells$end_climate, na.rm = TRUE)
sd.climate <- sd(icells$end_climate, na.rm = TRUE)

vocc_by_sp[[i]] <- vocc_by_sp[[i]] %>% 
  mutate(
    mean_climate = mean.climate, sd_climate = sd.climate,
    std_end_climate = (end_climate - mean.climate)/(sd.climate*2),
    std_end_climate2 = std_end_climate^2
    )

}else{

if (length(variable_names) == 2) {

vocc_by_sp[[i]] <- vocc_by_sp[[i]] %>% rename(
  source_end_climateA = X1, source_end_climateB = X2
  )

#browser()

icells <- vocc_by_sp[[i]] %>% distinct(icell, .keep_all = TRUE)
mean.climateA <- mean(icells$source_end_climateA, na.rm = TRUE)
sd.climateA <- sd(icells$source_end_climateA, na.rm = TRUE)
mean.climateB <- mean(icells$source_end_climateB, na.rm = TRUE)
sd.climateB <- sd(icells$source_end_climateB, na.rm = TRUE)

vocc_by_sp[[i]] <- vocc_by_sp[[i]] %>% 
  mutate(
    mean_climateA = mean.climateA, sd_climate = sd.climateA,
    std_end_climateA = (source_end_climateA - mean.climateA)/(sd.climateA*2),
    std_end_climateA2 = std_end_climateA^2,
    mean_climateB = mean.climateB, sd_climate = sd.climateB,
    std_end_climateB = (source_end_climateB - mean.climateB)/(sd.climateB*2),
    std_end_climateB2 = std_end_climateB^2
    )

#browser()
  end_climate <- purrr::map_df(strsplit(vocc_by_sp[[i]]$end_climate, " "), 
                   function(x) data.frame(t(as.numeric(x))))

  vocc_by_sp[[i]] <- cbind(vocc_by_sp[[i]], end_climate) %>%
    rename(end_climate_chr = end_climate, end_climate = X1, end_climateB = X2 )

} else {

 stop("Doesn't currently support testing vectors based on more than two climate variables at once.")
}
}

}
names(vocc_by_sp) <- names(biomass_list)

# glimpse(icells)
# range(icells$perc_change)
# range(vocc_by_sp[[i]]$std_change)
# range(vocc_by_sp[[i]]$X)

# glimpse(vocc_by_sp[[i]])
# trim spatial outliers 
# Xmax <- quantile(vocc_by_sp[[i]]$X, 0.9995)
# Xmin <- quantile(vocc_by_sp[[i]]$X, 0.005)
# std_max <- quantile(vocc_by_sp[[i]]$std_change, 0.9995)
# 
# vocc_by_sp[[i]] <- vocc_by_sp[[i]] %>% filter ( X >= Xmin & X <= Xmax ) 
```

@@

mature_plot <- ggplot(data, aes(x = end_climate, y = (end_density # - start_density
  ), colour = cell_type, #size = start_density, 
  group = cell_type)) +
  geom_jitter(width = 0.05, alpha = 0.45) + #, size = 2
  geom_smooth(method="lm", formula = y ~ x + I(x^2) + I(x^3)) +
  #   , aes(x = end_climate, y = (end_density #- start_density
  #  )), inherit.aes = FALSE) +
  scale_size_continuous(guide = "none") +
  scale_y_continuous(trans = fourth_root_power) +
  ylab(paste("Final biomass")) +
  xlab(paste("Final value of", climate,"for all cells")) +
  scale_colour_viridis_d(begin = 0.1, end = 0.5) + 
  gfplot::theme_pbs() + theme(legend.position = c(0.8, 0.8)) + 
  ggtitle(paste(survey, "mature", species))

mature_plot
```

@@

coefs <-cbind(var, coefs)
#coefs$var
#coefs$beta
#coefs$se
coefs$species <- spp 
coefs$lifestage <- "adult"
coefs$climate <- climate 
coefs$min_thresholds <- min_thresholds
coefs$max_thresholds <- max_thresholds
coefs$optimum <- optimal_value
coefs$start_time <- start_time
coefs$timespan <- delta_t_total
coefs$region <- region
coefs$ssid_string <- ssid_string

@@

coefs$species <- spp 
coefs$lifestage <- "immature"
coefs$climate <- climate 
coefs$min_thresholds <- min_thresholds
coefs$max_thresholds <- max_thresholds
coefs$optimum <- optimal_value
coefs$start_time <- start_time
coefs$timespan <- delta_t_total
coefs$region <- region
coefs$ssid_string <- ssid_string

>>>>

#climate <- "temperature"
temp_threshold <- c(0.75)
optimal_temp <- 6.5 # yellow-eye 4

@@

climate <- "do and temperature"
max_thresholds <- c(Inf, 0.75)
min_thresholds <- c(0.25, Inf)
critical_thresholds <- c(-0.25, 0.75)
optimal_values <- c(2, 6.5)
```

@@

if (climate == "temperature") {
  variable_names <- c("temp")
  min_thresholds <- c(Inf)
  delta_threshold <- temp_threshold
  optimal_values <- optimal_temp

@@

if (climate == "do") {
  variable_names <- c("do_est")
  max_thresholds <- c(Inf)
  delta_threshold <- -1*(do_threshold)
  optimal_values <- optimal_do

@@

optimal_string <- paste0(optimal_values, collapse = "-")
climate_string <- gsub(" ", "-", gsub("\\/", "-", tolower(climate)))
```

@@

vocc_plot <- gridExtra::grid.arrange(
  vocc_plot_temp, vocc_plot_do,
  ncol = 2,
  top = grid::textGrob(paste("do of ", optimal_values[1], "less", min_thresholds[1],"or temperature of ", optimal_values[2], "plus", max_thresholds[2], " over ", start_time, "-", start_time + delta_t_total, ""))
)

@@

# Extract biomass values to match spatial and temporal range of climate vectors
```{r}

# #   #climate <- "do and temperature"
# max_thresholds <- c(Inf, 0.5)
# min_thresholds <- c(0.1, Inf)
# # critical_thresholds <- c(-0.25, 0.75)
# optimal_values <- c(2, 6.5)

vectors <- trim_vector_data(vocc, variable_names, optimal_values, min_thresholds, max_thresholds)

if (length(vectors[ ,2]) == 0) {
  other_coefs <- readRDS("data/vocc-by-spp-sdmTMB-coefs.RDS")
  coefs <- other_coefs[1,]
  coefs$var <- "NA"
  coefs$beta <- "NA"
  coefs$se <- "NA"
  coefs$species <- spp 
  coefs$lifestage <- "NA"
  coefs$climate <- climate 
  coefs$min_thresholds <- min_string
  coefs$max_thresholds <- max_string
  coefs$optimum <- optimal_string
  coefs$start_time <- start_time
  coefs$timespan <- delta_t_total
  coefs$region <- region
  coefs$ssid_string <- ssid_string

  stop("No vectors. Conditions saved to 'data/vocc-byspp-sdmTMB-coefs'.")
}

#glimpse(vectors)
```

# Combine vector and biomass data

```{r}
vocc_by_sp <- lapply(biomass_list, function(i) {

biomass_change <- make_trend_data(i,
  ssid = model_ssid,
  start_time = start_time,
  skip_time = skip_time,
  end_time = end_time,
  input_cell_size = input_cell_size,
  scale_fac = scale_fac,
  delta_t_total = delta_t_total,
  delta_t_step = delta_t_step,
  indices = indices,
  variable_names = "est_exp"
)

vectors <- vectors %>%
      mutate(
        start_cell = icell,
        vector_id = paste0(id, "_to_", tid),
        distance2 = distance*distance,
        source_end_value = e, # may need to turn into numeric vector later...
        target_value = target_values
      )

@@

vector_source <- left_join(vectors, start, by = c( "icell", "x", "y")) %>% 
  mutate(cell_type = "source", end_climate = source_end_value, 
    source_density = start_density)

end <- biomass_change %>% 
  rename(start_density = est_exp_s, 
         end_density = est_exp_e, 
         target_X = x, target_Y = y) %>% 
  select(start_density, end_density, target_X, target_Y)

vector_target <- left_join(vectors, end, by = c("target_X", "target_Y")) %>% 
  mutate(cell_type = "target", end_climate = target_value)

@@

vector_target <- left_join(vector_target, source_density, by = c("x", "y"))

# combine source and target cell data
vocc_by_sp <- rbind(vector_source, vector_target) %>% 
  mutate(cell_type = as.factor(cell_type)) %>%
  filter(!is.na(icell)) %>%

  # remove potential source cells with lowest 1% quantile of fish biomass  
  filter(source_density > quantile(source_density, 0.01) ) %>%

  # calculate percent change in biomass
  mutate(perc_change = (end_density - start_density)/start_density)

end_climate <- purrr::map_df(strsplit(vocc_by_sp$end_climate, " "), 
                   function(x) data.frame(t(as.numeric(x))))

names(end_climate) <- paste0("end_", variable_names)

vocc_by_sp <- cbind(vocc_by_sp, end_climate) %>%
    rename(end_climate_chr = end_climate)

# get mean and sd for densities without cell duplication
icells <- vocc_by_sp %>% distinct(icell, .keep_all = TRUE)

mean.density <- mean(icells$source_density, na.rm = TRUE)
sd.density <- sd(icells$source_density, na.rm = TRUE)
mean.change <- mean(icells$perc_change, na.rm = TRUE)
sd.change <- sd(icells$perc_change, na.rm = TRUE)

@@

mean.var <- paste0("mean.", variable_names[j])
sd.var <- paste0("sd.", variable_names[j])

icells <- mutate(icells,
  (!! mean.var) := mean(UQ(rlang::sym(paste0(variable_names[j], "_e"))), na.rm = TRUE),
  (!! sd.var) := sd(UQ(rlang::sym(paste0(variable_names[j], "_e"))), na.rm = TRUE)
)
}

# calculate standardized density variables
vocc_by_sp <- vocc_by_sp %>% 
  mutate(
    std_source_density = (source_density - mean.density)/(sd.density*2), 
    mean_source_density = mean.density, sd_density = sd.density,
    mean_change = mean.change, sd_change = sd.change,
    std_change = (perc_change)/sd.change)

# calculate standardized climate variables
for (j in seq_along(variable_names)){

mean.var <- paste0("mean.", variable_names[j])
sd.var <- paste0("sd.", variable_names[j])
std.var <- paste0("std.", variable_names[j])
std.var2 <- paste0("std.", variable_names[j],"2")

vocc_by_sp <- mutate(vocc_by_sp,
  (!! mean.var) := icells[1, mean.var],
  (!! sd.var) := icells[1, sd.var],
  (!! std.var) := (UQ(rlang::sym(paste0(variable_names[j], "_e"))) - icells[1, mean.var])/ icells[1, sd.var],
  (!! std.var2) := ((UQ(rlang::sym(paste0(variable_names[j], "_e"))) - icells[1, mean.var])/ icells[1, sd.var])^2
  )
}

vocc_by_sp <- vocc_by_sp %>% 
  mutate(
    std_source_density = (source_density - mean.density)/(sd.density*2), 
    mean_source_density = mean.density, sd_density = sd.density,
    mean_change = mean.change, sd_change = sd.change,
    std_change = (perc_change)/sd.change)

# rename spatial coordinates for inclusion in models and plots without naming conflicts
vocc_by_sp <- vocc_by_sp %>% 
  rename(X = x, Y = y)

vocc_by_sp
})

# glimpse(vocc_by_sp)
# trim spatial outliers 
# Xmax <- quantile(vocc_by_sp[[i]]$X, 0.9995)
# Xmin <- quantile(vocc_by_sp[[i]]$X, 0.005)
# std_max <- quantile(vocc_by_sp[[i]]$std_change, 0.9995)
# 
# vocc_by_sp[[i]] <- vocc_by_sp[[i]] %>% filter ( X >= Xmin & X <= Xmax ) 
```

@@

#for (j in seq_along(variable_names)) {

mature_plot <- lapply(variable_names, function(j) { 
ggplot(data, aes(x = UQ(rlang::sym(paste0("end_", j))), y = (end_density # - start_density
  ), colour = cell_type, #size = start_density, 
  group = cell_type)) +
    geom_point() +
  #geom_jitter(width = 0.05, alpha = 0.45) + #, size = 2
  geom_smooth(method="lm", formula = y ~ x + I(x^2) + I(x^3)) +
  #   , aes(x = end_climate, y = (end_density #- start_density
  #  )), inherit.aes = FALSE) +
  scale_size_continuous(guide = "none") +
  scale_y_continuous(trans = fourth_root_power) +
  ylab(paste("Final biomass")) +
  xlab(paste("Final value of", variable_names,"for all cells")) +
  scale_colour_viridis_d(begin = 0.1, end = 0.5) + 
  gfplot::theme_pbs() + theme(legend.position = c(0.8, 0.8)) + 
  ggtitle(paste(survey, "mature", species))
})
print(mature_plot)
```

@@

coefs <-cbind(var, coefs)
#coefs$var
#coefs$beta
#coefs$se
coefs$species <- spp 
coefs$lifestage <- "adult"
coefs$climate <- climate 
coefs$min_thresholds <- min_thresholds
coefs$max_thresholds <- max_thresholds
coefs$optimum <- optimal_values
coefs$start_time <- start_time
coefs$timespan <- delta_t_total
coefs$region <- region
coefs$ssid_string <- ssid_string

@@

coefs$species <- spp 
coefs$lifestage <- "immature"
coefs$climate <- climate 
coefs$min_thresholds <- min_thresholds
coefs$max_thresholds <- max_thresholds
coefs$optimum <- optimal_values
coefs$start_time <- start_time
coefs$timespan <- delta_t_total
coefs$region <- region
coefs$ssid_string <- ssid_string
